Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-07 numerical-zeus.tex

\begin{eqnarray}
\rho \frac{\partial\vecv}{\partial t} &=& - \nabla p - \rho \nabla \phi 
- \nabla \cdot \textbf{Q} \\
\frac{\partial e}{\partial t} &=& -p \nabla \cdot \vecv - \textbf{Q} : \nabla \vecv, 
\end{eqnarray}
%where \textbf{v} is the baryon velocity, $\rho$ is the mass density, $p$ is the pressure, $e$ is the internal energy density of the gas and

@@

Finally, the third source step is the compression term and is given by
\begin{equation}
e^{n+c}_j = e^{n+b}_j \left( \frac{1 - (\Delta t/2) (\gamma - 1) (\nabla \cdot \vecv)_j }
                           {1 + (\Delta t/2) (\gamma - 1) (\nabla \cdot \vecv)_j } \right)
\end{equation}
We have used the notation $(\nabla \cdot \vecv)_j$ to indicate the (potentially) multi-dimensional velocity divergence evaluated at the cell center position $x_j$.  This equation differs from the previous ones in that in the multi-dimensional case, still only one finite difference equation is evaluated, but the divergence becomes multi-dimensional.

>>>>

\begin{eqnarray}
\rho \frac{\partial\vecv}{\partial t} &=& - \grad p - \rho \grad \phi 
- \div \textbf{Q} \\
\frac{\partial e}{\partial t} &=& -p \div \vecv - \textbf{Q} : \grad \vecv, 
\end{eqnarray}
%where \textbf{v} is the baryon velocity, $\rho$ is the mass density, $p$ is the pressure, $e$ is the internal energy density of the gas and

@@

Finally, the third source step is the compression term and is given by
\begin{equation}
e^{n+c}_j = e^{n+b}_j \left( \frac{1 - (\Delta t/2) (\gamma - 1) (\div \vecv)_j }
                           {1 + (\Delta t/2) (\gamma - 1) (\div \vecv)_j } \right)
\end{equation}
We have used the notation $(\div \vecv)_j$ to indicate the (potentially) multi-dimensional velocity divergence evaluated at the cell center position $x_j$.  This equation differs from the previous ones in that in the multi-dimensional case, still only one finite difference equation is evaluated, but the divergence becomes multi-dimensional.

### 2013-06-07 parallel.tex

At each complete root grid time step (or less often if specified),
each timer is then communicated to the root processor where it
calculates the mean, standard deviation, minimum, and maximum for each
of the timers across all processors.  For level timers, there are additional
attributes such as the number of cell updates, the current number of
grids, and the average cells updates per second per MPI process.  This
information is then 
output to a logfile.  This provides a simplified interface to the user
that can be used to diagnose performance issues as well as estimate a
given problem type's scalability.  In addition to the logfile, we have
developed a plotting interface for quickly producing figures that
process the data from the logfile.  These capabilities are described
in the online documentation, along with further discussion of the
performance measurement implementation.

@@

In this section, we develop a simple, approximate model to estimate
the computational time and memory required to complete a given
calculation.  Note that we are {\it not} trying to model parallel
performance (which was discussed in the previous section), but have an
even simpler goal: to determine how to scale computational time
estimates as we increase the simulation resolution.  This is a
straightforward thing to do in codes that use static grids, as the
computational effort per time step is constant.  However, for an AMR
calculation, the number of grids that will be generated during the run
is not known in advance, and so the CPU time per problem time can vary
drastically throughout the calculation.  For example, at the beginning
of the simulation when the densities are nearly uniform, only the
static grid is required and the calculation progresses rapidly.
However, as structure forms and dense clumps are generated, the number
of grid points swells by orders of magnitude (an increase of $10^3$ is
not uncommon) and most of the CPU time is consumed at late times.
Therefore, simply performing a few steps at the beginning of the
calculation does produce a good estimate of the required CPU time.

Nevertheless, we can try to determine how the compute time scales for
a given run as we increase the resolution.  First, we neglect
particles and concentrate on the time taken by the grids, which can be
justified both theoretically and empirically\footnote{Since the
potential is calculated on the grid, the only particle costs are:
depositing mass to the grid, interpolating accelerations, and updating
particle positions and velocities.  These are relatively
computationally inexpensive operations.}.  Second, we break the
problem down slightly, and examine the scaling over a short enough
period of time that the grid structure does not change significantly
(i.e. the number of grids at each level remains approximately
constant).  We then assume that the whole run scales in the same way,
or in other words, that changes in the resolution affect the
simulation in the same way at each time step.  This is usually a good
approximation, as the most costly parts of the calculation typically
don't change their characteristics substantially between timesteps.

To make progress, we assume that the computational cost to advance a
single cell by one timestep is a constant.  This can be incorrect if
the chemistry solver requires many iterations, but is usually fairly
accurate.  For a unigrid calculation, the time would by proportional
to $N_{\rm root}^4$ since the number of cells scales as $N_{\rm
root}^3$ and, assuming the Courant condition is the 
factor that controls the time step, the number of steps to advance the
calculation over a given time interval is
proportional to $N_{\rm root}.$ Therefore, to advance a hierarchy a
given time interval, we find, accounting for all levels and using a refinement
factor of 2,
\begin{equation} t_{\rm SU} = C_1 \sum_l f_l N_{\rm root}^4 2^{4l}
\end{equation} where $C_1$ is a constant, which can be thought of as
the time taken to advance a single cell.  The factor $f_l$ is the
fraction of the volume on a given level that is actually refined.  By
definition $f_0 = 1$, and $f_l \le 1$.  In writing down this equation,
we neglect a number of costs that are not directly proportional to
cell count, including communication between processors, the $\log{N}$
factor for the root grid FFT, cache misses, optimizations, and other
costs associated with processing the hierarchy.  The first item on
this list, in particular, is clearly important for large processor
counts; however, we neglect parallel considerations in this section.

>>>>

At each complete root grid timestep (or less often if specified),
each timer is then communicated to the root processor where it
calculates the mean, standard deviation, minimum, and maximum for each
of the timers across all processors.  For level timers, there are additional
attributes such as the number of cell updates, the current number of
grids, and the average cells updates per second per MPI process.  This
information is then 
output to a logfile.  This provides a simplified interface to the user
that can be used to diagnose performance issues as well as estimate a
given problem type's scalability.  In addition to the logfile, we have
developed a plotting interface for quickly producing figures that
process the data from the logfile.  These capabilities are described
in the online documentation, along with further discussion of the
performance measurement implementation.

@@

In this section, we develop a simple, approximate model to estimate
the computational time and memory required to complete a given
calculation.  Note that we are {\it not} trying to model parallel
performance (which was discussed in the previous section), but have an
even simpler goal: to determine how to scale computational time
estimates as we increase the simulation resolution.  This is a
straightforward thing to do in codes that use static grids, as the
computational effort per timestep is constant.  However, for an AMR
calculation, the number of grids that will be generated during the run
is not known in advance, and so the CPU time per problem time can vary
drastically throughout the calculation.  For example, at the beginning
of the simulation when the densities are nearly uniform, only the
static grid is required and the calculation progresses rapidly.
However, as structure forms and dense clumps are generated, the number
of grid points swells by orders of magnitude (an increase of $10^3$ is
not uncommon) and most of the CPU time is consumed at late times.
Therefore, simply performing a few steps at the beginning of the
calculation does produce a good estimate of the required CPU time.

Nevertheless, we can try to determine how the compute time scales for
a given run as we increase the resolution.  First, we neglect
particles and concentrate on the time taken by the grids, which can be
justified both theoretically and empirically\footnote{Since the
potential is calculated on the grid, the only particle costs are:
depositing mass to the grid, interpolating accelerations, and updating
particle positions and velocities.  These are relatively
computationally inexpensive operations.}.  Second, we break the
problem down slightly, and examine the scaling over a short enough
period of time that the grid structure does not change significantly
(i.e. the number of grids at each level remains approximately
constant).  We then assume that the whole run scales in the same way,
or in other words, that changes in the resolution affect the
simulation in the same way at each timestep.  This is usually a good
approximation, as the most costly parts of the calculation typically
don't change their characteristics substantially between timesteps.

To make progress, we assume that the computational cost to advance a
single cell by one timestep is a constant.  This can be incorrect if
the chemistry solver requires many iterations, but is usually fairly
accurate.  For a unigrid calculation, the time would by proportional
to $N_{\rm root}^4$ since the number of cells scales as $N_{\rm
root}^3$ and, assuming the Courant condition is the 
factor that controls the timestep, the number of steps to advance the
calculation over a given time interval is
proportional to $N_{\rm root}.$ Therefore, to advance a hierarchy a
given time interval, we find, accounting for all levels and using a refinement
factor of 2,
\begin{equation} t_{\rm SU} = C_1 \sum_l f_l N_{\rm root}^4 2^{4l}
\end{equation} where $C_1$ is a constant, which can be thought of as
the time taken to advance a single cell.  The factor $f_l$ is the
fraction of the volume on a given level that is actually refined.  By
definition $f_0 = 1$, and $f_l \le 1$.  In writing down this equation,
we neglect a number of costs that are not directly proportional to
cell count, including communication between processors, the $\log{N}$
factor for the root grid FFT, cache misses, optimizations, and other
costs associated with processing the hierarchy.  The first item on
this list, in particular, is clearly important for large processor
counts; however, we neglect parallel considerations in this section.

### 2013-06-07 table_of_variables_and_constants.tex

  \multicolumn{2}{l}{\underline{Heat Conduction}} \\[5pt]
  $\fcond$ & thermal conduction heat flux \\
  $\kappa_{sp}$ & Spitzer conduction coefficient \\
  $f_{Sp}$ & fraction of Spitzer conductivity \\[3pt]

>>>>

  \multicolumn{2}{l}{\underline{Heat Conduction}} \\[5pt]
  $\fcond$ & thermal conduction heat flux \\
  $\kappa_{\rm cond}$ & thermal conduction coefficient \\
  $\kappa_{\rm sp}$ & Spitzer conductivity \\
  $f_{\rm sp}$ & fraction of Spitzer conductivity \\[3pt]

### 2013-06-07 tests-sedov.tex

\noindent where $E_0$ is the initial energy injection, $\rho_0$ is the
background density and $\alpha = 1.0$ for cylindrical symmetry and an
ideal gas with $\gamma = 1.4$.  For the full derivation see
\citet{Sedov1959}.  Third, as the spherical shock expands, its
symmetry, or lack thereof, serves to highlight any directional
preferences of the hydrodynamics solvers.  The test presented here is
the two-dimensional 
version that is included in the \enzo\ distribution. The
three-dimensional results from this test, both for \enzo\ and three
other leading astrophysics codes, can be found in \citet{Tasker2008}.

>>>>

\noindent where $E_0$ is the initial energy injection, $\rho_0$ is the
background density and $\alpha = 1.0$ for cylindrical symmetry and an
ideal gas with $\gamma = 1.4$; for the full derivation see
\citet{Sedov1959}.  Third, as the spherical shock expands, its
symmetry, or lack thereof, serves to highlight any directional
preferences of the hydrodynamics solvers.  The test presented here is
the two-dimensional 
version that is included in the \enzo\ distribution. The
three-dimensional results from this test, both for \enzo\ and three
other leading astrophysics codes, can be found in \citet{Tasker2008}.

### 2013-06-07 tests-sodshocktube.tex

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/SodShockTube.eps}
\caption{The density distribution of the classic Sod Shock Tube for
three different solvers (from left to right column) and with (bottom
row) and without (top row) AMR.  In each case 100 zones are used on
the root level and the results are shown at $t=0.25$.  All cells are
plotted and color-coded by level (at the time shown, only small region
surrounding the contact discontinuity and the shock are refined).  In
each panel, we show the analytic solution as a solid line and the $E_1$
error norm in the upper right.}
\label{fig.sodshocktube}
\end{center}
\end{figure}

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/SodShockTube.eps}
\caption{The density distribution of the classic Sod Shock Tube for
three different solvers (from left to right column) and with (bottom
row) and without (top row) AMR.  In each case 100 zones are used on
the root level and the results are shown at $t=0.25$.  All cells are
plotted and color-coded by level with purple 
indicating level 0, green level 1 and red level 2 (at the time shown, only small region
surrounding the contact discontinuity and the shock are refined).  In
each panel, we show the analytic solution as a solid line and the $E_1$
error norm in the upper right.}
\label{fig.sodshocktube}
\end{center}
\end{figure}
